OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-05 main.tex
Added figure 5B and fixed small typos

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
viarations and enhancing batch effects. To address this issue, we
developed an alternative appoach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrantions, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
\textbf{Availability:} OneD is implemented as an R package available at
\href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}.\\	
textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

\begin{align*}
t_i &= \sum_j^n{x_{ij}} \sim  NB(\lambda_i, \theta), \\
\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)},
\end{align*}

@@

\begin{align}
\notag
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\label{eq:xhat}
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
\end{align}

@@

\begin{equation*}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
\end{equation*}

@@

We developed the routines contained in the \texttt{oned} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{lgf} and
\textit{caib} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occuring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Table
\ref{tab:samples} for details). Given a set of experiments and a metric,
we first computed all pairwise combinations between experiments and then
classified the comparisons according to the characteristics of each pair
(cell type, protocol, batch, and treatment).

@@

The principle of OneD is to explicitly model Hi-C biases on a single
variable: the total amount of contacts for each bin of the matrix (see
\ref{sec:model}). The reason for this choice of that the total amount of
contacts is approximately proportional to the local copy number. For
instance, a duplicated region in a diploid genome will show on average a
50\% increase in the number of contacts. Discontinuities of the amount of
contacts thus correspond to changes of the copy number.

@@

Other biases affect the total amount of contacts in a countinuous but not
necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the number of retriction enzyme sites
in T47D, a breast cancer cell line with aberrant and unstable karyotype.
Four clouds are visible. Each corresponds to a copy number between one and
four. In all of them, the relationship flattens as the number of
restriction sites increases. For this reason, OneD fits a non-linear
relationship between the total amount of contacts and the known sources of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated genome-wide and each cell of the matrix is then
corrected using equation (\ref{eq:xhat}). Note that the corrected amount
of contacts is still proportional to the copy number. The correction thus
represents a more stable and precise estimation of the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosme 17 of a T47D. OneD greatly reduces the wiggling of the total
amount of contacts.

@@

The \textit{caib} and \textit{ice} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}C and Supplementary Figure~3). This is an undesirable
effect, as it obscures the biological variability. In the same vein, these
methods decreased the similarity between samples that received the same
treatment (Figures~\ref{fig:aberrant}A), suggesting that the normalization
process is detrimental to the biological signal in these two cases. The
method \textit{vanilla} followed the same trend but to a lesser extent,
consistent with the fact that it consists of a single \textit{ice}
iteration.

@@

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}B and D).

All the methods, including the absence of normalizaton, succeeded in
recognizing the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging. OneD
increased the discrimination power compared to the raw matrices, but all
the other methods decreased it. Actually, they performed little better
than random on this task.

@@

In these conditions, variations of the protocol had a strong effect on the
impact of the differen normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caib} and \textit{ice} increased the similarity when
the protocols were different, but decreased it when the protocols were the
same. The effect was stronger when comparing identical cell types, but the
same trend appeared when compaing different cell types, indicating that
these methods may enhance or reduce biological variation, depending on the
context. Once more, \textit{vanilla} followed the same trend as
\textit{ice} to a lesser extent. The \textit{lgf} method increased the
similarity when comparing the same cells with different experimental
protocols, and had little to no effect in the other cases. This indicates
that \textit{lgf} is a safe choice in this case.

@@

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
pairs of biologicate pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
on this problem, even though it was not always above the others methods at
every position of the graph.

@@

Taken togetheer, these results indicate that OneD performs as well as the
best normalization methods even with normal karyotypes.

@@

Finally, we compared the computational speed of the different
normalization methods. One of the main reasons for the broad acceptance of
\textit{vanilla} and \textit{ICE} as standards is the speed of the
available implementation \citep{imakaev2012iterative}. This is even more
significant that current explicit methods \citep{servant2012hitc} are much
slower in comparison.

Unlike the other methods, OneD corrects a single variable instead of the
whole matrix. It uses only the total number of contacts per bin and thus
estimates the model parameters much faster than previous explicit methods.
We measured the running time of the different tools on a 3.3~GHz machine
with 62 GB~RAM, always using the default paramters. Figure\ref{fig:times}
shows the running times of the different methods on the samples described
in Table~\ref{tab:samples} at 100 kbp resolution. The fastest method is
\textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
span between the two.

OneD is the second fastest method and it always ran in less than 1 min in
the conditions of the benchmark. In all cases, the memory footprint was
less than 300 MB. Interestingly, the running time of OneD is much more
homogeneous than that of the other methods. The reason is that the size of
the regression problem to be solved by the \texttt{mgcv} package is always
the same at fixed reslution.

@@

\begin{figure}
\centerline{\includegraphics[width=.35\textwidth]{img/times_sina_figure5.pdf}}
\caption{
Computing time of the bias correction methods. Each dot corresponds to
one sample. The only method faster than ours under performs in all sample
comparisons. Note the logarithmic scale on the y-axis.}
\label{fig:times}
\end{figure}

@@

Here we introduced OneD, a fast computational method to normalize Hi-C
matrices. OneD was developed ground up to address the need to normalize
data collected on samples with aberrant karyotypes, but it applies
seemlessly to the case of normal karyotypes. We showed that OneD performs
significantly better than other methods when the cells present karyotypic
aberrations (Figure~\ref{fig:aberrant}), and that it performs equall well
as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
that OneD is on average faster than \textit{ice}, which makes it very
competitive from the point of view of computational speed.

@@

Depending on the intention of the user, the effect of the copy number
should either be kept or removed. This is why OneD leaves does not perform
the correction by default, but allows the user to obtain a
diploid-equivalent Hi-C map computed from a hidden Marko model. The
resulting matrices have a near constant amount of contacts per bin, but
the artifacts caused by the misfit between the genome of the sample and
the reference genome are still present (for instance, the artifacts caused
by large scale inversions are not changed in any way).

>>>>

\abstract{
  \textbf{Motivation:} The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented advance in our understanding of genome
structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but none of them addresses the common issue of
abnormal karyotypes. Many experiments are performed in cancer cell lines,
which typically harbor large-scale copy number variations that create
visible defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.\\
	\textbf{Results:}
We observed that current normalization methods perform badly in the
presence of large-scale copy number variations, obscuring biological
variations and enhancing batch effects. To address this issue, we
developed an alternative approach designed to take into account chromosomal
abnormalities. The method, called OneD, increases reproducibility among
replicates of Hi-C samples with abnormal karyotype, thus significantly
outperforming previous methods.  In the absence of chromosomal
aberrations, OneD fared equally well as state-of-the-art methods, making
it a safe choice for Hi-C normalization. OneD is fast and scales well in
terms of computing resources for resolutions up to 1 kbp.\\	
\textbf{Availability:} OneD is implemented as an R package available at
\href{http://www.github.com/qenvio/dryhic}
{http://www.github.com/qenvio/dryhic}.\\	
\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
\textbf{Supplementary information:} Supplementary data are available at
\textit{Bioinformatics}	online.}

@@

\begin{align*}
t_i = \sum_j^n{x_{ij}} &\sim  NB(\lambda_i, \theta) \\
\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)}
\end{align*}

@@

\begin{align}
\notag
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\label{eq:xhat}
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
\end{align}

@@

\begin{equation*}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}
\end{equation*}

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{lgf} and
\textit{caib} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}. Finally, the last option, proposed by
\cite{yan2017hicspector} is to measure the Pearson correlation between the
last eigenvectors of the Laplacian of the Hi-C matrix. This approach
borrows the concepts of spectral clustering \citep{von2007tutorial} and
amounts to comparing high level features of the matrix.

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Table
\ref{tab:samples} for details). Given a set of experiments and a metric,
we first computed all pairwise combinations between experiments and then
classified the comparisons according to the characteristics of each pair
(cell type, protocol, batch and treatment).

@@

The principle of OneD is to explicitly model Hi-C biases on a single
variable: the total amount of contacts for each bin of the matrix (see
\ref{sec:model}). The reason for this choice is that the total amount of
contacts is approximately proportional to the local copy number. For
instance, a duplicated region in a diploid genome will show on average a
50\% increase in the number of contacts. Discontinuities of the amount of
contacts thus correspond to changes of the copy number.

@@

Other biases affect the total amount of contacts in a continuous but not
necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
between the amount of contacts and the number of restriction enzyme sites
in T47D, a breast cancer cell line with aberrant and unstable karyotype.
Four clouds are visible. Each corresponds to a copy number between one and
four. In all of them, the relationship flattens as the number of
restriction sites increases. For this reason, OneD fits a non-linear
relationship between the total amount of contacts and the known sources of
bias (by default the G+C content, the number of restriction sites and the
mappability of the reads).

The biases are estimated genome-wide and each cell of the matrix is then
corrected using equation (\ref{eq:xhat}). Note that the corrected amount
of contacts is still proportional to the copy number. The correction thus
represents a more stable and precise estimation of the copy number.
Figure~\ref{fig:totals}B shows the corrected number of contacts along
chromosome 17 of a T47D. OneD greatly reduces the wiggling of the total
amount of contacts.

@@

The \textit{caib} and \textit{ice} methods increased the similarity
between the different cell lines (Figures~\ref{fig:aberrant}A and
\ref{fig:aberrant}B and Supplementary Figure~3). This is an undesirable
effect, as it obscures the biological variability. In the same vein, these
methods decreased the similarity between samples that received the same
treatment (Figure~\ref{fig:aberrant}A), suggesting that the normalization
process is detrimental to the biological signal in these two cases. The
method \textit{vanilla} followed the same trend but to a lesser extent,
consistent with the fact that it consists of a single \textit{ice}
iteration.

@@

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D).

All the methods, including the absence of normalization, succeeded in
recognizing the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging. OneD
increased the discrimination power compared to the raw matrices, but all
the other methods decreased it. Actually, they performed little better
than random on this task.

@@

In these conditions, variations of the protocol had a strong effect on the
impact of the different normalization methods (Figure~\ref{fig:normal}A).
For instance, \textit{caib} and \textit{ice} increased the similarity when
the protocols were different, but decreased it when the protocols were the
same. The effect was stronger when comparing identical cell types, but the
same trend appeared when comparing different cell types, indicating that
these methods may enhance or reduce biological variation, depending on the
context. Once more, \textit{vanilla} followed the same trend as
\textit{ice} to a lesser extent. The \textit{lgf} method increased the
similarity when comparing the same cells with different experimental
protocols, and had little to no effect in the other cases. This indicates
that \textit{lgf} is a safe choice in this case.

@@

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
 biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
on this problem, even though it was not always above the others methods at
every position of the graph.

@@

Taken together, these results indicate that OneD performs as well as the
best normalization methods even with normal karyotypes.

@@

Finally, we compared the computational speed of the different
normalization methods. One of the main reasons for the broad acceptance of
\textit{vanilla} and \textit{ICE} as standards is the speed of the
available implementation \citep{imakaev2012iterative}. This is even more
significant as current explicit methods \citep{servant2012hitc} are much
slower in comparison.

Unlike the other methods, OneD corrects a single variable instead of the
whole matrix. It uses only the total number of contacts per bin and thus
estimates the model parameters much faster than previous explicit methods.
We measured the running time of the different tools on a 3.3~GHz machine
with 62 GB~RAM, always using the default parameters. Figure~\ref{fig:times}A
shows the running times of the different methods on the samples described
in Table~\ref{tab:samples} at 100 kbp resolution. The fastest method is
\textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
span between the two. OneD also scales well in terms of bin size, as can
be seen in Figure~\ref{fig:times}B, where the different corrections were
applied to a reduced part of the genome at increasing resolutions.

OneD is the second fastest method and it always ran in less than 1 min in
the conditions of the benchmark. In all cases, the memory footprint was
less than 300 MB. Interestingly, the running time of OneD is much more
homogeneous than that of the other methods. The reason is that the size of
the regression problem to be solved by the \texttt{mgcv} package is always
the same at fixed resolution

@@

\begin{figure}
\centerline{\includegraphics[width=.5\textwidth]{img/times_figure5.pdf}}
\caption{
Computing time of the bias correction methods. A. Total time for the entire genome at 100 kbp resolution. Each dot corresponds to
one sample. The only method faster than ours under performs in all sample
comparisons. B. Time to correct a reduced genome (chr19-22) of one sample at different resolutions. Note the logarithmic scale on the y-axis on both panels.}
\label{fig:times}
\end{figure}

@@

Here we introduced OneD, a fast computational method to normalize Hi-C
matrices. OneD was developed ground up to address the need to normalize
data collected on samples with aberrant karyotypes, but it applies
seamlessly to the case of normal karyotypes. We showed that OneD performs
significantly better than other methods when the cells present karyotypic
aberrations (Figure~\ref{fig:aberrant}), and that it performs equally well
as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
that OneD is on average faster than \textit{ice}, which makes it very
competitive from the point of view of computational speed.

@@

Depending on the intention of the user, the effect of the copy number
should either be kept or removed. This is why OneD does not perform
the correction by default, but allows the user to obtain a
diploid-equivalent Hi-C map computed from a hidden Markov model. The
resulting matrices have a near constant amount of contacts per bin, but
the artifacts caused by the misfit between the genome of the sample and
the reference genome are still present (for instance, the artifacts caused
by large scale inversions are not changed in any way).
